---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
```


<LAQ1>

# 1. Build a random forest model that can predict cell-of-origin (COO) using the training/testing cohort (70/30 training/testing split). Tune the model to select an appropriate number of trees and try value to produce the lowest out of bag(OOB) error. Report the error of the model in the trining and testing cohort(OOB, type 1, type 2 errors and a ROC curve). Discuss how well your model works on these data.

```{r}
library(Biobase)
library(randomForest)
library(caret)
library(pROC)
library(tidyverse)

cohort2 <- readRDS("/Users/emily/Desktop/For_LAQ1_required_data/DLBCL_cohort_2/DLBCL.cohort.2.rds")
set.seed(123)
ind <- sample(1:2, ncol(cohort2), replace = TRUE, prob = c(0.7,0.3))
train <- cohort2[,ind==1]
test <- cohort2[,ind==2]
train.rf <- cbind(COO = as.factor(train$COO),
                 t(exprs(train)) %>% as.data.frame())
test.rf <- cbind(COO = as.factor(test$COO),
                t(exprs(test)) %>% as.data.frame())
tune <- tuneRF(train.rf[,-1], train.rf$COO,
               ntreeTry=100,
               stepFactor=1.5,
               improve=0.01)
best.mtry <- tune[which.min(tune[,2]),1]
rf <- randomForest(train.rf[,-1], train.rf$COO,
                   ntree=200,
                   mtry=best.mtry)
print(rf)
```
```{r}
train.pred <- predict(rf, train.rf)
confusionMatrix(train.pred, train.rf$COO)
train.prob <- predict(rf, train.rf, type="prob")
roc.train <- roc(train.rf$COO, train.prob[,2])
plot(roc.train)
test.pred <- predict(rf, test.rf)
confusionMatrix(test.pred, test.rf$COO)
test.prob <- predict(rf, test.rf, type="prob")
roc.test <- roc(test.rf$COO, test.prob[,2])
plot(roc.test)
```

The optimal mtry value was 192, which gave the lowest OOB error during tuning (~1.99%). The final model had an OOB error of 4.38%. From the confusion matrix, the type 1 error was 7.02% and the type 2 error was 2.19%. This shows the model performs slightly better for GCB than ABC. The sensitivity was 0.9623 and specificity was 1.0000, with a balanced accuracy of 0.9811. 
The ROC curve shows an AUC close to 1, indicating very strong classification performance.

# 2. Select the 1000 most important features in the model and use these to recreate the model - Report the error of this model in the training and testing cohort(OOB, type 1, type 2 errors & ROC). Discuss whether it performs better or worse than the previous model.

```{R}
imp <- importance(rf) %>% as.data.frame()
top1000 <- rownames(imp %>% arrange(-MeanDecreaseGini))[1:1000]
train.rf.1000 <- train.rf[, c("COO", top1000)]
test.rf.1000 <- test.rf[, c("COO", top1000)]
rf.1000 <- randomForest(train.rf.1000[,-1], train.rf.1000$COO,
                        ntree=200, mtry=sqrt(1000))
print(rf.1000)
```

This model reduced the OOB error to 1.2%, compared to 4.38% in the previous model. The type 1 error decreased to 1.75% and the type 2 error decreased to 0.73%, showing improved performance for both classes. 
The model performs better with fewer features, suggesting that removing less informative genes improves classification accuracy.

# 3. Recursively retain the model using the n most important features (where n is a reducing number of features) to find the optimum number of genes to include in the training. Report the error of the model in the training and testing cohort, including OBB, type 1, type 2 errors & ROC. Discuss the n value chosen, explaining your strategy for gene selection and describe which of the 3 models you feel best models the data.

```{R}
feature_sizes <- c(1000, 500, 200, 100, 50, 20)
results <- data.frame()
for (n in feature_sizes) {
  genes <- rownames(imp %>% arrange(-MeanDecreaseGini))[1:n]
  train.sub <- train.rf[, c("COO", genes)]
  test.sub <- test.rf[, c("COO", genes)]
  model <- randomForest(train.sub[,-1], train.sub$COO,
                        ntree=150, mtry=sqrt(n))
  test.pred <- predict(model, test.sub)
  cm <- confusionMatrix(test.pred, test.sub$COO)
  acc <- cm$overall["Accuracy"]
  results <- rbind(results, data.frame(n=n, accuracy=acc))}
results
```

The highest accuracy (0.9899) was observed at both 200 and 50 genes.50 genes were selected as the optimal number to reduce model complexity while maintaining accuracy.When fewer than 50 genes were used, accuracy went down, which means using too few genes can remove important information.

# 4. Using your final optimum model, predict the cell of origin in Cohort 1 (the cohort used in the practical). Report the type 1 and type 2 errors and aROC curve in this validation cohort. Discuss whether this model works within this validation cohort and explain your reasoning.

```{R}
cohort1 <- Biobase::combine(
  readRDS("/Users/emily/Desktop/For_LAQ1_required_data/DLBCL_cohort_1/DLBCL.eSet.1.rds"),
  readRDS("/Users/emily/Desktop/For_LAQ1_required_data/DLBCL_cohort_1/DLBCL.eSet.2.rds"),
  readRDS("/Users/emily/Desktop/For_LAQ1_required_data/DLBCL_cohort_1/DLBCL.eSet.3.rds"),
  readRDS("/Users/emily/Desktop/For_LAQ1_required_data/DLBCL_cohort_1/DLBCL.eSet.4.rds"))

final.genes <- rownames(imp %>% arrange(-MeanDecreaseGini))[1:100]
cohort1.rf <- cbind(COO=as.factor(cohort1$COO),
                    t(exprs(cohort1)) %>% as.data.frame())
cohort1.sub <- cohort1.rf[, c("COO", final.genes)]

pred <- predict(model, cohort1.sub)
confusionMatrix(pred, cohort1.sub$COO)
prob <- predict(model, cohort1.sub, type="prob")
roc.val <- roc(cohort1.sub$COO, prob[,2])
plot(roc.val)
```
The sensitivity of this model was 0.8941 and specificity was 0.7974, with a balanced accuracy of 0.8458. Compared to the training and testing results, the performance decreased, indicating that the model does not generalise perfectly to independent data. 
The ROC curve shows an AUC of approximately 0.88–0.90, suggesting that the model still has good predictive ability. 
The model performs reasonably well in the validation cohort, but the drop in performance suggests some overfitting during training.




<LAQ 2>

# 1.Perform the unsupervised clustering analysis using non-negative matrix factorisation(NMF). Find the most appropriate number of clusters, and provide all the essential statistics and plots to support your cluster.
```{R}
library(NMF)
exp_data <- read.table("/Users/emily/Desktop/For_LAQ2_required_data/exp_data_ESCC_90samples.txt", header=TRUE, row.names=1)
mads <- apply(exp_data,1,mad)
d_exp <- exp_data[rev(order(mads))[1:1500],]
estim.r <- nmf(d_exp, 2:6, nrun=10, seed=123456)
plot(estim.r)
```

```{r}
consensusmap(estim.r)
res <- nmf(d_exp, 4, "brunet", seed=123456)
nmf_cluster <- predict(res)
table(nmf_cluster)
```
The consensus map at k=4 shows clearer cluster structure, while higher k values appear more unstable. At k=4, the cophenetic coefficient shows a local increase before decreasing again, and the silhouette score also improves slightly compared to k=3. I selected k=4 as the most appropriate number of clusters, and the samples were divided into four groups (21, 33, 13, and 23), suggesting a reasonable clustering structure.

# 2. Based on your selection of number of clusters, select and report the most significant gene signatures for each cluster (i.e., top 20 genes based on the differential expression of each cluster against all others.)
```{R}
library(limma)
cl <- nmf_cluster
limma_res <- list()
for (i in 1:4) {
  coi <- d_exp[, cl==i]
  rc  <- d_exp[, cl!=i]
  factors <- factor(rep(c("coi","others"),
                 c(ncol(coi), ncol(rc))))
  mm <- model.matrix(~factors-0)
  colnames(mm) <- levels(factors)
  fit <- lmFit(cbind(coi, rc), mm)
  contrast <- makeContrasts(coi-others, levels=mm)
  fit2 <- contrasts.fit(fit, contrast)
  fit2 <- eBayes(fit2)
  deg <- topTable(fit2, number=20)
  limma_res[[i]] <- deg
  print(paste("Cluster", i))
  print(head(deg, 20))}
```
Cluster 1 : VIM, CD74, HLA-A/C/DRA, and IFI30.
Cluster 2 : KRT6A, KRT17, KRT14 and SLC2A1.
Cluster 3 : CSTB, CD24, ANXA1, and S100A9.
Cluster 4 : G6PD, NQO1, TXN, and TKT.

# 3. Perform GSEA and report the most unregulated pathways for each cluster, and comment on your findings.
```{R}
install.packages("BiocManager")
BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("enrichplot")
BiocManager::install("DOSE")
library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r}
gene_lists <- list()
for (i in 1:4) {
  coi <- d_exp[, cl == i]
  rc  <- d_exp[, cl != i]
  factors <- factor(rep(c("coi","others"),
                 c(ncol(coi), ncol(rc))))
  mm <- model.matrix(~factors - 0)
  colnames(mm) <- levels(factors)
  fit <- lmFit(cbind(coi, rc), mm)
  contrast <- makeContrasts(coi - others, levels = mm)
  fit2 <- contrasts.fit(fit, contrast)
  fit2 <- eBayes(fit2)
  res <- topTable(fit2, number = Inf)
  gene_list <- res$logFC
  names(gene_list) <- rownames(res)
  gene_list <- gene_list[!is.na(gene_list)]
  gene_list <- sort(gene_list, decreasing = TRUE)
  gene_lists[[i]] <- gene_list}

GSEA_res <- list()
for (i in 1:4) {
  gse <- gseGO(
    geneList = gene_lists[[i]],
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pvalueCutoff = 0.05)
  GSEA_res[[i]] <- gse}
```

```{r}
for (i in 1:4) {
  print(paste("Cluster", i))
  print(head(as.data.frame(GSEA_res[[i]]), 10))}
```

Cluster 1 : cell migration, cell adhesion, and locomotion-related pathways
Cluster 2 : response to stimulus, defense response, and immune-related pathways
Cluster 3 : keratinization and epithelial/epidermal cell differentiation pathways
Cluster 4 : xenobiotic metabolism, response to toxic substances, and metabolic processes

Each cluster shows distinct biological functions, suggesting that the clustering successfully separates biologically meaningful groups.

# 4. Repeat the procedure 1-3 above using the consensus clustering algorithm(e.g., ConsensusClusterPlus R package) and report all the required findgings.
```{R}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ConsensusClusterPlus")
library(ConsensusClusterPlus)

consensus_res <- ConsensusClusterPlus(
  as.matrix(d_exp),
  maxK=6,
  reps=50,
  pItem=0.8,
  pFeature=1,
  clusterAlg="hc",
  distance="pearson",
  seed=123)
cc_cluster <- consensus_res[[4]]$consensusClass
table(cc_cluster)
```
```{r}
library(limma)
group <- factor(cc_cluster)
design <- model.matrix(~0 + group)
colnames(design) <- paste0("C", levels(group))
fit <- lmFit(d_exp, design)
contrast.matrix <- makeContrasts(
  C1vsAll = C1 - (C2+C3+C4)/3,
  C2vsAll = C2 - (C1+C3+C4)/3,
  C3vsAll = C3 - (C1+C2+C4)/3,
  C4vsAll = C4 - (C1+C2+C3)/3,
  levels = design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
limma_cc <- list()
top20_genes <- list()
for(i in 1:4){
  res <- topTable(fit2, coef=i, number=Inf)
  limma_cc[[i]] <- res
  top20_genes[[i]] <- rownames(res)[1:20]}
top20_genes

library(clusterProfiler)
library(org.Hs.eg.db)
gene_lists <- list()
for (i in 1:4) {
  gene_list <- limma_cc[[i]]$logFC
  names(gene_list) <- rownames(limma_cc[[i]])
  gene_list <- sort(gene_list, decreasing = TRUE)
  gene_lists[[i]] <- gene_list}
GSEA_cc <- list()
for (i in 1:4) {
  gse <- gseGO(
    geneList = gene_lists[[i]],
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pvalueCutoff = 0.05)
  GSEA_cc[[i]] <- gse}
for (i in 1:4) {
  print(paste("Cluster", i))
  print(head(as.data.frame(GSEA_cc[[i]])))}
```
Cluster 1 is dominated by keratin genes, indicating an epithelial/keratinization state. Cluster 2 shows strong immune and inflammatory signals. Cluster 3 is characterized by WNT/TWIST genes, suggesting an EMT-like phenotype. Cluster 4 shows epithelial and metabolic features.
GSEA : cluster 1 is linked to migration/adhesion, cluster 2 to immune response, cluster 3 to differentiation, and cluster 4 to metabolic processes.
Overall, the clusters are stable and biologically distinct.


# 5. Finally, compare the clusters and cluster membership between that based on NMF and that from consensus clustering. Are these clusters similar? Are the unregulated and down regulated pathways identified from the two algorithms similar? Comment on the comparison.
```{R}
table(NMF = nmf_cluster, Consensus = cc_cluster)
```
Most samples from NMF clusters 2, 3 and 4 were all grouped into Consensus cluster 1. The two methods do not give similar clustering results. 
NMF separates the samples into four distinct groups, while consensus clustering puts them into one large cluster, so it fails to clearly distinguish subtypes.
The biological differences identified by NMF are likely more meaningful.

# 6. Can you suggest the most appropriate procedure for any unsupervised clustering analysis of a given dataset? How to generate the most robust, reliable clusters?

Unsupervised clustering shouldn’t depend on just one method. It’s better to first select variable genes to reduce noise. Then try different clustering methods and different numbers of clusters. The main point is to choose clusters that are stable, not just based on one result. After that, check if the clusters make biological sense using methods like differential expression or GSEA. Robust clustering means the results are consistent and easy to interpret biologically. 



